Continuous Deductive Reasoning over Streamed RDF Data

### 2015-11-03 article.tex
Major reshuffle of TA-* section, bringing justification ahead of examples, and defining satisfiability of datalog rules by fact instances.  Also neatened up markup (mostly using \prime instead of ')

A conventional Datalog program $\Pi = \langle R, F\rangle$ comprises a
set of clauses, divided into a finite set of rules $R$ and a finite
set of ground facts $F$. Facts are formulae of first-order logic, and
rules are of the form:
\[ \alpha_0 \Leftarrow \alpha_1 \land \ldots \land \alpha_n \]
where $\alpha_i$ are atomic formulae. Note that we have excluded any
universally quantified variables for clarity.

@@

We define a {\em justification} of a set of facts that are entailed by
a Datalog program as a minimal subset of the program that is
sufficient for the entailment to hold. More precisely, for a Datalog
program $\Pi = \langle R, F \rangle$ and a set of entailed facts $E$
where $\langle R, F \rangle \models E$, the program $\langle J_R, J_F
\rangle$ with $J_R \subseteq R$ and $J_F \subseteq F$ is a
justification for $E$ iff $\langle J_R, J_F \rangle \models E$ and,
for all $J'_R \subset J_R$ or $J'_F \subset J_F$, $\langle J'_R, J'_F
\rangle \not\models E$.

We simplify this notion of justification by making the assumption that
a justification can be defined in terms of a minimal subset of the
ground facts of the program; the rules of the program are therefore
implicitly considered to be part of every justification. Any
subsequent references to {\em justification} are to this simplified
sense, which is defined as follows: given a Datalog program $\Pi =
\langle R, F \rangle$ and a set of entailed facts $E$ where $\langle
R, F \rangle \models E$, the set of facts $J \subseteq F$ is a
justification for $E$ iff $\langle R, J \rangle \models E$ and, for
all $J' \subset J$, $\langle R, J' \rangle \not\models E$. Note that a
given set of entailed facts may not have a unique justification.

@@

When considering a TA-Datalog program, in which there may be many
distinct instances of a given fact, temporal information regarding
when instances of that fact became true would be lost if such a
program were to entail sets of plain facts without temporal
annotations. We therefore extend the model theory for Datalog from
\refdef{continuous datalog: datalog program} by considering the
Herbrand base of all fact instances $\hat{H}^T$ (all facts with all
instants). As before, a fact instance $e$ is entailed by a program
$\Pi^T$ iff each interpretation satisfying every clause in $\Pi^T$
also satisfies $e$, written $\Pi^T \models e$. The {\em entailment of the
TA-Datalog program} $\Pi^T$ is therefore:
\[ cons^T(\Pi^T) = \{ e \in \hat{H}^T \mid \Pi^T \models e \} \]

Informally, the set of instants appearing in some set $E^T$ of fact
instances entailed by some program ${\langle R, F^T \rangle}$ will be
a subset of the instants appearing in the fact instances in $F^T$.  In
addition, the instant $t$ in a fact instance $\langle e, t \rangle$
that is entailed by a TA-Datalog program $\Pi^T = \langle R^T, F^T
\rangle$ will be bounded from above by the maximum instant in any fact
instance in $F^T$. A TA-Datalog program may entail fact instances that
are only distinguished by their instants (and are therefore considered
distinct from each other, as per \refdef{continuous datalog: fact
  instance}).

Consider the following example TA-Datalog program:
\[
\Pi^T = \langle \{ c \Leftarrow a \land b \}, \{ \langle a, 1 \rangle, \langle a, 2 \rangle, \langle b, 1 \rangle, \langle b, 2 \rangle  \} \rangle
\]

The entailment of the program $\Pi^T$ will be:
\[cons^T(\Pi^T) = \{\langle c, 1 \rangle, \langle c, 2 \rangle\}\]

It follows that applying the ${strip()}$ function to all the fact
instances in both the set of fact instances $F^T$ in some ${\Pi^T =
  \langle R, F^T \rangle}$ and the entailment of the program $\Pi^T$
(the set of fact instances $cons^T(\Pi^T)$) maintains the relationship
of entailment; if $\langle R, F^T\rangle \models \langle e, t
\rangle$, then $\langle R, strip(F^T) \rangle \models strip(\langle e,
t \rangle)$. This relationship is shown in \reffig{TD-D}.

Reconsidering the example TA-Datalog program above, we may reduce it
to the following equivalent conventional Datalog program $\Pi$:
\begin{align*}
\Pi &= \langle \{ c \Leftarrow a \land b \}, strip(\{ \langle a, 1 \rangle, \langle a, 2 \rangle, \langle b, 1 \rangle, \langle b, 2 \rangle \}) \rangle \\
    &= \langle \{ c \Leftarrow a \land b \}, \{ a, b \} \rangle
\end{align*}
Trivially, the entailment of this program $\Pi$ will therefore be:
\begin{align*}
cons(\Pi) &= strip(\{\langle c, 1 \rangle, \langle c, 2 \rangle\}) \\
 &= \{ c \}  
\end{align*}
\end{definition}

@@

For an entailed fact instance $\langle e, t\rangle$ and a
justification of that fact instance $J^T$, the instant $t$ at which
$\langle e, t \rangle$ becomes true is taken to be the latest instant
at which any of the fact instances in the justification became true:
\[ t = max(\{ t' \mid \langle j', t' \rangle \in J^T \}) \]

Note that, although an entailed fact instance $\langle e, t\rangle$
may have multiple justifications, all the justifications for $\langle
e, t \rangle$ will share the same maximum instant; a justification
with a different maximum instant $t' \neq t$ would entail a different
fact instance $\langle e, t' \rangle$. Considering again the example
TA-Datalog program in \refdef{continuous datalog: ta datalog program},
the entailed fact instance $\langle c, 2 \rangle$ would have the
following set of justifications:
\[
  \{ \ \{ \langle a, 1 \rangle, \langle b, 2 \rangle \},  \{ \langle a, 2 \rangle, \langle b, 1 \rangle \},  \{ \langle a, 2 \rangle, \langle b, 2 \rangle \} \  \}
\]

@@

>>>>

A conventional Datalog program $\Pi = \langle R, F\rangle$ comprises a
set of clauses, divided into a finite set of rules $R$ and a finite
set of ground facts $F$. Facts are atomic formulae of first-order logic, and
rules are of the form:
\[ \alpha_0 \Leftarrow \alpha_1 \land \ldots \land \alpha_n \]
where $\alpha_i$ are atomic formulae. Note that we have excluded any
universally quantified variables for clarity.

@@

We define a {\em justification} of a set of facts that are entailed by
a Datalog program as a minimal subset of the program that is
sufficient for the entailment to hold. More precisely, for a Datalog
program $\Pi = \langle R, F \rangle$ and a set of entailed facts $E$
where $\langle R, F \rangle \models E$, the program $\langle J_R, J_F
\rangle$ with ${J_R \subseteq R}$ and ${J_F \subseteq F}$ is a
justification for $E$ iff $\langle J_R, J_F \rangle \models E$ and,
for all ${J^\prime_R \subset J_R}$ or ${J^\prime_F \subset J_F}$, ${\langle J^\prime_R, J^\prime_F
\rangle \not\models E}$.

We simplify this notion of justification by making the assumption that
a justification can be defined in terms of a minimal subset of the
ground facts of the program; the rules of the program are therefore
implicitly considered to be part of every justification. Any
subsequent references to {\em justification} are to this simplified
sense, which is defined as follows: given a Datalog program $\Pi =
\langle R, F \rangle$ and a set of entailed facts $E$ where $\langle
R, F \rangle \models E$, the set of facts $J \subseteq F$ is a
justification for $E$ iff $\langle R, J \rangle \models E$ and, for
all $J^\prime \subset J$, $\langle R, J^\prime \rangle \not\models E$. Note that a
given set of entailed facts may not have a unique justification.

@@

When considering a TA-Datalog program, in which there may be many
distinct instances of a given fact, temporal information regarding
when instances of that fact became true would be lost if such a
program were to entail sets of plain facts without temporal
annotations. We therefore extend the model theory for Datalog from
\refdef{continuous datalog: datalog program} by considering the
Herbrand base of all fact instances $\hat{H}^T$ (all facts with all
instants). As before, a fact instance $\langle e, t_e \rangle$ is 
entailed by a program $\Pi^T$ iff each interpretation
$\mathcal{I}^T \subseteq \hat{H}^T$ satisfying every
clause (rule in $R$ and fact instance in $F^T$) in $\Pi^T$ also satisfies
$\langle e, t_e \rangle$, written $\Pi^T \models \langle e, t_e \rangle$.
As with facts in positive Datalog, we say that an interpretation
$\mathcal{I}^T$ satisfies a fact instance $\langle f, t \rangle \in F^T$
iff $\langle f, t \rangle \in \mathcal{I}^T$.  We say the interpretation
$\mathcal{I}^T$ satisfies a rule
$(r = \alpha_0 \Leftarrow \alpha_1 \land \ldots \land \alpha_n) \in R$ iff,
given $\{\langle a_1, t_1 \rangle, \dots, \langle a_n, t_n \rangle\} \subseteq \mathcal{I}^T$
such that $\{a_1, \dots, a_n\}$ satisfies the body of $r$,
$\langle a_0, t_0 \rangle \in \mathcal{I}^T$ (where $t_0 = max(t_1,\dots,t_n)$,
see \refdef{continuous datalog: instance justification}) such that
$\{a_0, \dots, a_n\}$ satisfies $r$.  The {\em entailment of the TA-Datalog
program} $\Pi^T$ is therefore:
\[ cons^T(\Pi^T) = \{ \langle e, t_e \rangle \in \hat{H}^T \mid
\Pi^T \models \langle e, t_e \rangle \} \]

@@

For an entailed fact instance $\langle e, t_e\rangle$ and a
justification of that fact instance $J^T$, the instant $t_e$ at which
$\langle e, t_e \rangle$ becomes true is, intuitively, taken to be the
latest instant at which any of the fact instances in the justification
became true:
\[ t_e = max(\{ t \mid \langle j, t \rangle \in J^T \}) \]

Note that, although an entailed fact instance $\langle e, t_e\rangle$
may have multiple justifications, all the justifications for $\langle
e, t_e \rangle$ will share the same maximum instant; a justification
with a different maximum instant $t^\prime \neq t_e$ would entail a different
fact instance $\langle e, t^\prime \rangle$.

@@

From this definition of entailment in TA-Datalog, the set of instants
appearing in some set $E^T$ of fact instances entailed by some program
${\langle R, F^T \rangle}$ will be a subset of the instants appearing
in the fact instances in $F^T$.

Consider the following example TA-Datalog program:
\begin{equation}
  \Pi^T = \langle \{ c \Leftarrow a \land b \}, \{ \langle a, 1 \rangle, \langle a, 2 \rangle, \langle b, 1 \rangle, \langle b, 2 \rangle  \} \rangle
\end{equation}

The entailment of the program $\Pi^T$ will be:
\begin{equation}
  cons^T(\Pi^T) = \{\langle c, 1 \rangle, \langle c, 2 \rangle\}
\end{equation}
with the following justifications:
\begin{equation}
  \{ \ \{ \langle a, 1 \rangle, \langle b, 2 \rangle \},  \{ \langle a, 2 \rangle, \langle b, 1 \rangle \},  \{ \langle a, 2 \rangle, \langle b, 2 \rangle \} \  \}
\end{equation}

It follows that applying the ${strip()}$ function to all the fact
instances in both the set of fact instances $F^T$ in some ${\Pi^T =
  \langle R, F^T \rangle}$ and the entailment of the program $\Pi^T$
(the set of fact instances $cons^T(\Pi^T)$) maintains the relationship
of entailment; if $\langle R, F^T\rangle \models \langle e, t_e
\rangle$, then $\langle R, strip(F^T) \rangle \models strip(\langle e,
t_e \rangle)$. This relationship is shown in \reffig{TD-D}.

Reconsidering the example TA-Datalog program above, we may reduce it
to the following equivalent conventional Datalog program $\Pi$:
\begin{align*}
\Pi &= \langle \{ c \Leftarrow a \land b \}, strip(\{ \langle a, 1 \rangle, \langle a, 2 \rangle, \langle b, 1 \rangle, \langle b, 2 \rangle \}) \rangle \\
    &= \langle \{ c \Leftarrow a \land b \}, \{ a, b \} \rangle
\end{align*}
Trivially, the entailment of this program $\Pi$ will therefore be:
\begin{align*}
cons(\Pi) &= strip(\{\langle c, 1 \rangle, \langle c, 2 \rangle\}) \\
 &= \{ c \}  
\end{align*}
